---
title: |
 | Introduction to Data Analysis with R
 | Chapter 2. Introduction to R
author: Alfonso Zamora Saiz, Carlos Quesada González, Lluís Hurtado Gil and Diego
  Mondéjar Ruiz
output:
  html_document:
    toc: true
    toc_depth: 3
---

Each chunk represents a code window in the book. Please note that some chunks could not run properly or take a too much time to do it. The first line of these chunks is a comment, starting with the simbol `#` to clarify why is this.

# 2.1 Programming Languages and R

## 2.1.1 Some remarks on the history of programming languages

## 2.1.2 Paradigms and structure

## 2.1.3 The R language

## 2.1.4 RStudio

```{r}
3 + 4
```

## 2.1.5 First Steps

```{r}
for (i in 1 : 3) { 
  print(2 * i)
}
```

```{r}
temperatures <- c(18, 24, 26, 27, 19, 20, 22)
names(temperatures) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
```

```{r, eval=FALSE}
install.packages("ggplot2")
```

```{r}
library(ggplot2)
```

```{r, eval=FALSE}
installed.packages()
```

```{r, eval=FALSE}
update.packages(ask=FALSE)
```

```{r}
# Script for printing a sentence and a number
sentence <- "The price is" # defines the sentence 
price <- 249.99 # defines the price 
print(paste(sentence, price)) # prints both
```

Script for printing a sentence and a number
```{r}
sentence <- "The price is" # defines the sentence 
price <- 249.99 # defines the price 
print(paste(sentence, price)) # prints both
```

```{r}
help(matrix) 
?matrix
```

```{r}
help(package="ggplot2")
```

```{r}
help(diamonds, package="ggplot2")
```

```{r, eval=FALSE}
demo(graphics)
```

# 2.2 Data Structures

## 2.2.1 Variables and basic arithmetic

```{r}
3
```

```{r}
3 + 4
```

```{r}
3 - 5
```

```{r}
2 * 6
```

```{r}
5 / 3
```

```{r}
15 %% 2
```

```{r}
15 %/% 2
```

```{r}
x <- 4
```

```{r}
x
```

```{r}
y <- 5
```

```{r}
z <- x + y 
z
```

```{r}
number.decimal <- 4.3
class(number.decimal)
```

```{r}
2.73e4
```

```{r}
1 / 2.73e4
```

```{r}
word <- "Hello"
class(word)
```

```{r}
yes.no <- TRUE
class(yes.no)
```

```{r}
TRUE + TRUE
```

```{r}
TRUE + FALSE
```

```{r}
TRUE * FALSE
```

```{r}
FALSE * FALSE
```

```{r}
null.object <- NULL
null.object
```

```{r}
missing <- NA
missing
```

```{r}
1 + NA
```

```{r}
TRUE - NA
```

## 2.2.2 Vectors

```{r}
vector.numbers <- c(1, 2, 3, 4)
vector.numbers
```

```{r}
length(vector.numbers)
```

```{r}
vector.characters <- c("Today", "Yesterday", "Tomorrow")
vector.characters
```

```{r}
vector.logicals <- c(TRUE , FALSE)
vector.logicals
```

```{r}
c(NA, 3, 4)
```

```{r}
4 : 12
```

```{r}
rep(c(3, -1, 0.5), times=3)
```

```{r}
rep(c(3, -1, 0.5), each=3)
```

```{r}
rep(c(3, -1, 0.5), times=c(2, 1, 3))
```

```{r}
rep(c(3, -1, 0.5), length.out=8)
```

```{r}
seq(0, 10, 2.5)
```

```{r}
temperatures <- c(28, 29, 27, 27, 30)
names(temperatures) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
temperatures
```

```{r}
rains <- c(0, 5, 6, 0, 2)
names(rains) <- names(temperatures)
```

```{r}
days <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(rains) <- days
rains
```

```{r}
temperatures + 273.15
```

```{r}
temperatures * 1.8 + 32
```

```{r}
rains / (temperatures + 273.15)
```

```{r}
total.rains <- sum(rains)
total.rains
```

```{r}
rains[2]
```

```{r}
rains.beginning.week <- rains[c(1, 2)]
rains.beginning.week
```

```{r}
temperature.midweek <- temperatures[2 : 4]
temperature.midweek
```

```{r}
temperature.thursday <- temperatures["Thursday"]
temperature.thursday
```

```{r}
rains.beginning.week <- rains[c("Monday", "Tuesday")]
rains.beginning.week
```

```{r}
6 > 3
```

```{r}
4 == 5
```

```{r}
rains > 0
```

```{r}
3 == 4 & 3 == 3
```

```{r}
3 == 4 | 3 == 3
```

```{r}
3 < 5 & 4 > 2
```

```{r}
TRUE & TRUE & FALSE
```

```{r}
xor(3 > 2, FALSE)
```

```{r}
xor(TRUE , 4 == 4)
```

```{r}
not.rainy.days <- rains == 0
not.rainy.days
```

```{r}
hot.days <- temperatures >= 29
hot.days
```

```{r}
rains[hot.days]
```

```{r}
all(temperatures >= 28)
```

```{r}
any(temperatures == 30)
```

```{r}
which(temperatures == 27)
```

```{r}
some.vector <- c(3, 7, 9, 6, 2, 8)
order(some.vector)
```

```{r}
some.vector[order(some.vector)]
```

```{r}
sum(temperatures)
```

```{r}
max(temperatures)
```

```{r}
min(temperatures)
```

```{r}
range(temperatures)
```

```{r}
which(temperatures == max(temperatures))
```

## 2.2.3 Matrices

```{r}
matrix(data=NA, nrow=1, ncol=1, byrow=FALSE, dimnames=NULL)
```

```{r}
matrix(1 : 9, byrow=TRUE, nrow=3)
```

```{r}
matrix(1 : 15, nrow=2)
```

```{r}
matrix(c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"), nrow=3, byrow=TRUE)
```

```{r}
matrix(c(TRUE, TRUE, FALSE, TRUE), ncol=2)
```

```{r}
climate <- matrix(c(temperatures , rains), byrow=TRUE, nrow=2)
rownames(climate) <- c("Temperatures", "Rains")
colnames(climate) <- days
climate
```

```{r}
dim(climate)
```

```{r}
Winds <- c(30, 25, 22, 24, 18)
rbind(climate, Winds)
```

```{r}
total.climate <- rbind(climate, Winds)
totals <- rowSums(climate)
cbind(climate, totals)
```

```{r}
total.climate[2, 3]
```

```{r}
total.climate[, 4]
```

```{r}
total.climate[1, ]
```

```{r}
total.climate[, "Friday"]
```

```{r}
total.climate["Rains", ]
```

```{r}
ave.temperature <- mean(total.climate[1, ])
ave.temperature
```

```{r}
climate * 3
```

```{r}
climate ^ 2
```

```{r}
climate[2, ] / climate[1, ]
```

## 2.2.4 Factors

```{r}
sizes <- c("Small", "Big", "Big", "Medium", "Medium", "Small", "Medium", "Small", "Small")
sizes
```

```{r}
summary(sizes)
```

```{r}
factor.sizes <- factor(sizes)
factor.sizes
```

```{r}
summary(factor.sizes)
```

```{r}
levels(factor.sizes)
```

```{r}
my.colors <- c("Orange", "Red", "Red", "Yellow", "Blue", "Green")
factor.colors <- factor(my.colors)
factor.colors
```

```{r}
sizes2 <- c("Small", "Big", "Big", "Medium", "Medium", "Small", "Medium", "Small", "Small")
factor.sizes2 <- factor(sizes , ordered=TRUE, levels=c("Small", "Medium", "Big"))
factor.sizes2
```

```{r}
survey.vector <- c("M", "F", "F", "M", "M", "F", "M", "M")
factor.survey.vector <- factor(survey.vector)
factor.survey.vector
```

```{r}
summary(factor.survey.vector)
```

```{r}
levels(factor.survey.vector) <- c("Female", "Male")
factor.survey.vector
```

```{r}
factor.survey.vector[1] < factor.survey.vector[2]
```

```{r}
small <- factor.sizes2[1]
medium <- factor.sizes2[4]
big <- factor.sizes2[2]
small < medium
```

```{r}
small < big
```

```{r}
medium > big
```

```{r}
factor.sizes2[3] < factor.sizes2[5]
```

## 2.2.5 Data Frames

```{r}
?datasets
library(help="datasets")
```

```{r}
data()
```

```{r}
? OrchardSprays
OrchardSprays
```

```{r}
head(OrchardSprays)
```

```{r}
str(OrchardSprays)
```

```{r}
name <- c("Alfonso", "Carlos", "Lluis", "Diego")
last.name <- c("Zamora", "Quesada", "Hurtado", "Mondejar")
second.last.name <- c("Saiz", "Gonzalez", "Gil", "Ruiz")
age <- c(33, 32, 30, 37)
phd <- c("math", "math", "physics", "math")
office <- c(4, 14, 6, 8)
from.madrid <- c(FALSE, TRUE, FALSE, TRUE)
professors <- data.frame(name, last.name, second.last.name, age, phd, office, from.madrid)
str(professors)
```

```{r}
professors[2, 3]
```

```{r}
professors[1, ]
```

```{r}
professors[, 2]
```

```{r}
professors[1 : 2, ]
```

```{r}
professors$second.last.name
```

```{r}
professors$office[3]
```

```{r}
madrileans <- professors$from.madrid
professors[madrileans, ]
```

```{r}
subset(professors, subset=age < 31)
```

```{r}
subset(professors, subset=phd == "math")
```

```{r}
positions <- order(professors$age)
professors[positions, ]
```

```{r}
mydata.as.matrix <- matrix(1 : 9, nrow=3)
mydata.as.matrix
```

```{r}
mydata.as.data.frame <- as.data.frame(mydata.as.matrix)
mydata.as.data.frame
```

```{r}
mydata.as.data.frame$V1
```

```{r}
new.list <- list(days , factor.sizes , climate)
new.list
```

```{r}
new.list[1]
```

```{r}
new.list[[1]][2]
```

```{r}
new.list[[3]][1, 3 : 4]
```

```{r}
new.list <- list(the.days=days, the.factors=factor.sizes, the.data=climate)
new.list
```

```{r}
new.list$the.factors
```

```{r}
new.list$the.data[2, 5]
```

```{r}
new.list2 <- c(new.list , today="October 4")
str(new.list2)
```

# 2.3 Control Structures

## 2.3.1 Conditionals

```{r, eval=FALSE}
# Syntax, it does not run
if ("condition is satisfied") {
  "do something"
}
```

```{r}
x <- 3
if (x > 0) {
  print("Positive")
}
```

```{r}
x <- -5
if(x > 0){
  print("Positive")
}
```

```{r, eval=FALSE}
# Syntax, it does not run
if ("condition is satisfied"){
  "do something"
} else {
  "otherwise do something else"
}
```

```{r}
x <- -8
if (x > 0) {
  print("Positive")
} else {
  print("Negative")
}
```

```{r}
x <- 6
if (x > 0) {
  print("Positive")
} else {
  print("Negative")
}
```

```{r}
x <- 0
if (x > 0) {
  print("Positive")
} else {
  print("Negative")
}
```

```{r}
x <- 0
if (x > 0) {
  print("Positive")
} else if (x < 0) {
  print("Negative")
} else {
  print("Zero")
}
```

```{r}
x <- 9
ifelse(x > 0, "Positive", "Negative")
```

```{r}
(1 : 10) < 5
```

```{r}
ifelse((1 : 10) < 5, "Fail", "Pass")
```

## 2.3.2 Loops

```{r, eval=FALSE}
# Syntax, it does not run
for ("counter" in "vector of indices") {
  "do something"
}
```

```{r}
for (i in 1 : 10) {
  print(i)
}
```

```{r}
i
```

```{r}
v <- c()
v
```

```{r}
s <- 0
for (i in 1 : 10) {
  s <- s + i
  v[i] <- s
}
v
```

```{r}
odd <- 2 * (1 : 10) - 1
for (i in odd) {
  print(i)
}
```

```{r}
for (i in c(TRUE , FALSE)) {
print(TRUE == i)
}
```

```{r}
for (i in 1 : 10) {
  if (i < 5) {
    print("Fail")
  } else {
    print("Pass")
  }
}
```

```{r, eval=FALSE}
# Syntax, it does not run
while ("condition holds") {
  "do something"
}
```

```{r}
i <- 0
while (i < 10) {
  print(c(i,"is less than 10"))
  i <- i + 1
}
```

```{r}
i
```

```{r, eval=FALSE}
# infinite loop, do not run!!
j <- 1
while (j < 5) {
  print(c(j, "is less than 5"))
}
```

# 2.4 Functions and Operators

```{r}
mean(c(2, 5, 7))
```

```{r}
mean(matrix(1 : 4))
```

```{r}
sum(matrix(1 : 4))
```

```{r}
sqrt(2)
```

```{r}
log(34)
```

```{r}
exp(0)
```

```{r}
cos(0)
```

```{r}
is.logical(3 != 3)
```

```{r}
matrix(3 : 8, nrow=3)
```

```{r}
funtion.name <- function(argument1, argument2,...) {
  "body function"
}
```

```{r}
f <- function(x) {
  x / (1 - x) # the output is the evaluation of last line
}
```

```{r}
f <- function(x) x / (1 - x)
```

```{r}
f <- function(x) {
  return(x / (1 - x))
}
```

```{r}
f(2)
```

```{r}
y <- 4
f(y)
```

```{r}
f(2 * y)
```

```{r}
g <- function(x,y) {
  x ^ 2 - y / 5
}
```

```{r}
g(1, 2)
```

```{r}
g(2, 1)
```

```{r}
exp(1 : 10)
f(c(1, 4, 7))
```

```{r}
funtion.name <- function(name.argument1=default.value1, name.argument2=default.value2,...) {
  "body function"
}
```

```{r}
mat.vec <- function(a, b=2, flag=FALSE){
  if (flag) {
    matrix(1 : a, nrow=b)
  } else {
    1 : a
  }
}
```

```{r}
mat.vec(5)
```

```{r}
mat.vec(6, 3, TRUE)
```

```{r}
mat.vec(flag=TRUE , a=6, b=3)
```

```{r}
mat.vec(a=7, b=4)
```

```{r}
mat.vec(a=7, flag=TRUE)
```

```{r}
vector.selector <- function(v, pos=1){
  if (is.vector(v)) {
    return(v[pos])
  } else {
    stop("Variable v is not a vector")
  }
}
```

```{r}
vector.selector(3 : 8)
```

```{r}
vector.selector(c(1, 4, 5, 8, 2, 4, 3, 5), pos=6)
```

```{r}
# Error
# vector.selector(matrix(1 : 9))
```

```{r}
items <- function(x) list(len=length(x),total=sum(x),
mean=mean(x))
```

```{r}
data <- 1 : 10
result <- items(data)
result
```

```{r}
names(result)
```

```{r}
result$len
```

```{r}
result$tot
```

```{r}
result$mean
```

```{r}
log
```

```{r}
mean
```

```{r}
salutation <- function(x) print("Hello")
# Note that this output does not depend on the value of x
output <- sapply(1 : 5, salutation)
```

```{r}
output <- sapply(1 : 5, function(x) print("Hello"))
```

```{r}
lapply(cars ,mean)
```

```{r}
sapply(cars , mean)
```

## 2.4.1 Exercises

```{r}
g <- function(x, y) x ^ 2 - y / 5
```

