---
title: |
 | Introduction to Data Analysis with R
 | Chapter 5. Data Analysis with R
author: Alfonso Zamora Saiz, Carlos Quesada González, Lluís Hurtado Gil and Diego Mondéjar Ruiz
output:
  html_document:
    toc: true
    toc_depth: 3
---

Each chunk represents a code window in the book. Please note that some chunks could not run properly or take a too much time to do it. The first line of these chunks is a comment, starting with the simbol `#` to clarify why is this.

# 5.1 Descriptive Statistics

```{r}
library(mlbench)
data(PimaIndiansDiabetes2)
? PimaIndiansDiabetes2
head(PimaIndiansDiabetes2)
```

```{r}
attach(PimaIndiansDiabetes2)
```

## 5.1.1 Position measures

```{r}
mean(c(4, 5, 8, 2, 8, 4, 5, 9, 2, 7))
```

```{r}
mean(mass, na.rm=TRUE)
```

```{r}
sapply(PimaIndiansDiabetes2[, -9], mean, na.rm=TRUE)
```

```{r}
sales.per <- c(0.10, -0.05, 0.20, 0.05, -0.03)
sales <- 1 + sales.per
5000 * prod(sales)
```

```{r}
5000 * (1.050151) ^ 5
```

```{r}
5000 * (1.054) ^ 5
```

```{r}
library(psych)
geometric.mean(sales) - 1
```

```{r}
23 / 10 + 23 / 30
```

```{r}
46 / 20
```

```{r}
speeds = c(10, 30)
harmonic.mean(speeds, na.rm=TRUE)
```

```{r}
46 / 15
```

```{r}
median(1 : 7)
```

```{r}
median(1 : 6)
```

```{r}
median(mass , na.rm=TRUE)
```

```{r}
mean(c(1 : 7, 30))
```

```{r}
median(c(1 : 7, 30))
```

```{r}
sapply(PimaIndiansDiabetes2[, -9], median, na.rm=TRUE)
```

```{r}
library(DescTools)
Mode(mass)
```

```{r}
sapply(PimaIndiansDiabetes2, Mode, na.rm=TRUE)
```

```{r}
# plot histogram
hist(mass , col="seashell", border="black", xlab="Body mass index", main="Pima Indians Diabetes")
# add mean
abline(v=mean(mass, na.rm=TRUE), col="red", lwd=2)
# add median
abline(v=median(mass, na.rm=TRUE), col="blue", lwd=2)
abline(v=Mode(mass), col="seagreen", lwd=2) # add mode
legend(x="topright", c("Mean", "Median", "Mode"),
col=c("red", "blue", "seagreen"), lwd=c(2, 2)) # legend
```

```{r}
quantile(mass, probs=0.65, na.rm=TRUE)
```

```{r}
quantile(mass, na.rm=TRUE)
```

```{r}
quantile(mass, probs=seq(0, 1, 0.10), na.rm=TRUE)
```

```{r}
percent.fun <- ecdf(mass)
percent.fun(40)
```

```{r}
percent.fun(median(mass , na.rm=TRUE))
```

```{r}
summary(mass)
```

```{r}
summary(PimaIndiansDiabetes2)
```

## 5.1.2 Dispersion measures

```{r}
IQR(mass, na.rm=TRUE)
```

```{r}
sapply(PimaIndiansDiabetes2[, -9], IQR, na.rm=TRUE)
```

```{r}
boxplot(mass, xlab="Body mass index", col="papayawhip")
boxplot(pressure, xlab="Systolic blood pressure", col="pink")
```

```{r}
boxplot(mass)$out
```

```{r}
boxplot(pressure)$out
```

```{r}
boxplot(mass, range=3, xlab="Body mass index", col="papayawhip")
boxplot(pressure, range=3, xlab="Systolic blood pressure", col="pink")
```

```{r}
boxplot(mass , range=3)$out
```

```{r}
boxplot(pressure , range=3)$out
```

```{r}
var.p <- function(x){
  var(x, na.rm=TRUE) * (length(x) - 1) / (length(x))
}
sd.p <- function(x){
  sqrt(var.p(x))
}
```

```{r}
class1 <- c(5, 5, 5, 5, 5, 5, 5, 5, 5, 5)
class2 <- c(0, 0, 0, 0, 0, 10, 10, 10, 10, 10)
```

```{r}
var.p(class1)
```

```{r}
var.p(class2)
```

```{r}
sd.p(class1)
```

```{r}
sd.p(class2)
```

```{r}
var.p(mass)
```

```{r}
sd.p(mass)
```

```{r}
sapply(PimaIndiansDiabetes2[, -9], sd.p)
```

```{r}
sd.p(mass) / mean(mass , na.rm=TRUE)
```

```{r}
sd.p(pressure) / mean(pressure , na.rm=TRUE)
```

## 5.1.3 Shape measures

```{r}
library(moments)
```

```{r}
skewness(pressure, na.rm=TRUE)
```

```{r}
skewness(pregnant, na.rm=TRUE)
```

```{r}
x <- c(rep(1, 100), rep(2, 140), rep(3, 25), rep(4, 15), rep(5, 10))
y <- c(rep(1, 100), rep(2, 140), rep(3, 30), rep(4, 20), rep(5, 15))
skewness(x)
```

```{r}
skewness(y)
```

```{r}
x <- c(rep(1, 30), rep(2, 40), rep(3, 250), rep(4, 400), rep(5, 150), rep(6, 120), rep(7, 10))
skewness(x)
```

```{r}
kurtosis(triceps, na.rm=TRUE)
```

```{r}
sapply(PimaIndiansDiabetes2[,-9], FUN=kurtosis, na.rm=TRUE)
```

### 5.1.4 Concentration measures

```{r}
salary1 <- c(10, 15, 20, 21, 22, 35, 35, 38, 60, 100)
salary2 <- c(21, 22, 23, 25, 32, 38, 42, 46, 46, 50)
```

```{r}
Gini(salary1)
```

```{r}
Gini(salary2)
```

```{r}
Gini(pregnant, na.rm=TRUE)
```

```{r}
Gini(pressure, na.rm=TRUE)
```

```{r}
plot(Lc(pregnant), col="red", lwd=2, main="Lorentz curve for pregnant", xlab="Cumulative individuals", ylab="Cumulative values")
plot(Lc(pressure, na.rm=TRUE), col="blue", lwd=2, main="Lorentz curve for pressure", xlab="Cumulative individuals", ylab="Cumulative values")
```

```{r}
library(ineq)
Theil(pregnant)
```

```{r}
Theil(pressure)
```

## 5.1.5 Exercises

# 5.2 Statistical Inference

## 5.2.1 Random variables

```{r}
class(mass)
class(pregnant)
```

```{r}
sort(unique(pregnant))
```

```{r}
loaded <- data.frame(values=c(1, 2, 3, 4, 5, 6), probs=c(1 / 10, 1 / 10, 3 / 10, 2 / 10, 2 / 10, 1 / 10))
library(ggplot2)
ggplot(data=loaded , aes(x=values, y=probs)) +
  geom_bar(stat="identity", fill="aquamarine3") +
  ylim(c(0, 0.35))
```

```{r}
ggplot(data=PimaIndiansDiabetes2, aes(x=mass)) +
  geom_density() +
  xlab("Mass") +
  ylab("Density")
ggplot(data=PimaIndiansDiabetes2, aes(x=mass)) +
  stat_ecdf() +
  xlab("Mass") +
  ylab("Cumulative density")
```

```{r}
pdf.density <- density(mass , na.rm=TRUE)
```

```{r}
pdf.density$x[150]
```

```{r}
pdf.density$y[150]
```

```{r}
ecdf.cumul <- ecdf(mass)
ecdf.cumul(35)
```

```{r}
set.seed(5)
x <- 1 : 50
sample(x, size=10)
```

```{r}
set.seed(5)
sample(c("H", "T"), 25, replace=TRUE)
```

```{r}
set.seed(5)
sample(1 : 6, size=10, prob=c(0.1, 0.1, 0.1, 0.1, 0.1, 0.5), replace=TRUE)
```

## 5.2.2 Probability distributions

```{r}
? Distributions
```

```{r}
dbinom(2, size=5, prob=0.2)
```

```{r}
pbinom(2, 5, 0.2, lower.tail=TRUE)
```

```{r}
pbinom(2, 5, 0.2, lower.tail=FALSE)
```

```{r}
qbinom(0.95, 5, 0.2)
```

```{r}
set.seed(5)
rbinom(10, 5, 0.2)
```

```{r}
values <- 0 : 5
plot(values, dbinom(values , 5, 0.2), type="b", xlab="Values", ylab="Probabilities", main="Mass")
plot(values, pbinom(values , 5, 0.2), type="b", xlab="Values", ylab="Probabilities", main="Cumulative")
```

```{r}
values <- 0 : 11
plot(values, dpois(values, 5), type="b", xlab = "Values", ylab="Probabilities", main="Mass")
plot(ppois(values, 5), type="b", xlab = "Values", ylab="Probabilities", main="Cumulative")
```

```{r}
pnorm(80, 70, 15) - pnorm(60, 70, 15)
```

```{r}
x <- seq(40, 100, by=0.5)
curve(dnorm(x, 70, 15), xlim=c(40, 100), col="blue", lwd=2, xlab="x", ylab="f(x)", main="Density function N(70,15)")
curve(pnorm(x, 70, 15), xlim=c(40, 100), col="blue", lwd=2, xlab="x", ylab="f(x)", main="Cumulative function N(70,15)")
```

```{r}
set.seed(1)
sample <- rnorm(1000, 70, 15)
hist(sample, freq=FALSE, breaks=seq(20, 130, 10), col="seashell", xlab="Heart rate", main="Histogram for simulation of N(70,15)")
curve(dnorm(x, 70, 15), xlim=c(20, 130), col="blue", lwd=2, add=TRUE)
```

```{r}
qqnorm(mass, main="Normal Q-Q plot for mass")
qqline(mass, col="blue", lwd=2) # adds a reference line
```

```{r}
qqnorm(rnorm(1000, 70, 15), main="Normal Q-Q plot for N(70,15)")
qqline(rnorm(1000, 70, 15), col="blue", lwd=2)
```

```{r}
m = 10000 # sample size of the final variable
lambda = 4 # poisson parameter
par(mfrow=c(2, 2))
for (n in c(1, 5, 10, 100)) {
  x <- NULL #initialization of x
  for (i in 1:m) {
    x <- c(x, mean(rpois(n, lambda)))
  }
  x.normalized <- (x-lambda)/(sqrt(lambda/n))
  hist(x.normalized , col="seashell",
  main=paste("Mean of", n,"Po(4)"), xlab="Value")
}
```

```{r}
n = 100
m = 10000
mu = mean(mass , na.rm=TRUE)
sigma = sd(mass , na.rm=TRUE)
x <- NULL
for (i in 1:m) {
  x <- c(x, mean(sample(na.omit(mass), n)))
}
y = (x-mu)/(sigma/sqrt(n))
hist(y, col="seashell", main=paste(paste("Mean of 100",sep=" "),"mass",sep=" "), xlab="Value")
```

## 5.2.3 Confidence intervals and hypothesis tests

```{r}
mean(triceps, na.rm=TRUE)
```

```{r}
sd(triceps, na.rm=TRUE) ^ 2
```

```{r}
conf.level <- 0.95
t <- qt((1 - conf.level) / 2, df=length(triceps) - 1, lower.tail = FALSE)
sigma <- sd(triceps , na.rm=TRUE)
mean <- mean(triceps , na.rm=TRUE)
sd.mean <- sigma / sqrt(length(triceps) - sum(is.na(triceps)))
interval <- c(mean - t * sd.mean, mean + t * sd.mean)
interval
```

```{r}
MeanCI(triceps, na.rm=TRUE, conf.level=0.95)
```

```{r}
library(boot)
set.seed(1)
resampling <- boot(na.omit(triceps), function(x, i) mean(x[i]), R=10000)
boot.ci(resampling, conf = 0.95)
```

```{r}
VarCI(mass, na.rm=TRUE, conf.level=0.95)
```

```{r}
t.test(pressure, mu=70, alternative="greater")
```

```{r}
t.test(pressure, mu=72, alternative="two.sided")
```

```{r}
t.test(pressure, mu=71.4, alternative="two.sided")
```

```{r}
test <- t.test(pressure, mu=70, alternative="greater")
test$p.value
```

```{r}
t.test(insulin, glucose, mu=0, alternative="two.sided")
```

```{r}
t.test(insulin, glucose, mu=0, alternative="less")
```

```{r}
t.test(insulin, glucose, mu=30, alternative="two.sided")
```

```{r}
VarTest(pressure, sigma.squared=153)
```

```{r}
var.test(pressure, mass, ratio=3, alternative= "two.sided")
```

```{r}
library(snpar)
runs.test(c(1, 1, 1, 1, 1, 1, 1, 1, 0, 0))
```

```{r}
set.seed(5)
sample.vector <- rnorm(100000, 0, 1)
binary.vector <- ifelse(sample.vector < 0, 0, 1)
runs.test(binary.vector)
```

```{r}
ks.test(glucose, pressure)
```

```{r}
set.seed(5)
ks.test(rpois(100, 5), rpois(200, 5))
```

```{r}
shapiro.test(insulin)
```

```{r}
set.seed(5)
shapiro.test(rnorm(5000, 0, 1))
```

### 5.2.4 Exercises

## 5.3 Multivariate Statistics

### 5.3.1 Correlation and bivariate statistics

```{r}
ggplot(PimaIndiansDiabetes2, aes(x=mass, y=pressure)) +
  geom_point()
ggplot(PimaIndiansDiabetes2, aes(x=mass, y=triceps)) +
  geom_point()
```

```{r}
ggplot(PimaIndiansDiabetes2, aes(x=mass, y=triceps)) +
  geom_point() +
  geom_vline(xintercept=mean(mass, na.rm=TRUE), colour="red", size=1.2) +
  geom_hline(yintercept=mean(triceps, na.rm=TRUE), colour="blue", size=1.2) +
  annotate("text", label="mean of mass", angle=90, x=34, y=75, size=4, colour="red") +
  annotate("text", label="mean of triceps", x=55, y=25, size=4, colour="blue")
```

```{r}
cov(mass, pressure, use="pairwise.complete.obs")
```

```{r}
cov(mass, triceps, use="pairwise.complete.obs")
```

```{r}
cor(mass, pressure, use="pairwise.complete.obs")
```

```{r}
cor(mass, triceps, use="pairwise.complete.obs")
```

```{r}
cor(PimaIndiansDiabetes2[, -9], use="pairwise.complete.obs")
```

```{r}
library(corrplot)
corrplot(cor(PimaIndiansDiabetes2[,-9], use="pairwise.complete.obs"))
```

```{r}
plot(PimaIndiansDiabetes2)
```

## 5.3.2 Linear Model

```{r}
model <- lm(triceps ~ mass, data=PimaIndiansDiabetes2)
model$coefficients
```

```{r}
summary(model)
```

```{r}
summary(model)$r.squared
```

```{r}
ggplot(PimaIndiansDiabetes2, aes(x=mass, y=triceps)) +
  geom_point() +
  geom_smooth(method="lm", se=TRUE)
```

```{r}
predict(model, data.frame(mass=30), interval="confidence")
```

```{r}
library(psych)
pairs.panels(PimaIndiansDiabetes2[, -9],
method="pearson", hist.col="seashell",
density=TRUE , lm=TRUE)
```

## 5.3.3 Multivariable Linear Models

```{r}
model <- lm(triceps ~ mass + glucose + pressure)
summary(model)
```

```{r}
model <- lm(mass ~ ., data=PimaIndiansDiabetes2)
summary(model)
```

```{r}
contrasts(diabetes)
```

```{r}
model <- lm(mass ~ . -glucose - pedigree - age, data=PimaIndiansDiabetes2)
summary(model)
```

## 5.3.4 Nonlinear transformations

```{r}
linear.price <- lm(price ~ x, data=diamonds)
summary(linear.price)$adj.r.squared
```

```{r}
ggplot(diamonds, aes(y=price , x=x)) +
  geom_point(alpha=.5) +
  geom_smooth(method="lm") +
  xlim(3, 11) +
  ylim(0, 19000)
```

```{r}
poly.price <- lm(price ~ poly(x, 2), data=diamonds)
summary(poly.price)$adj.r.squared
```

```{r}
ggplot(diamonds , aes(y=price , x=x)) +
  geom_point(alpha=.5) +
  geom_smooth(method="lm", formula=y ~ poly(x, 2)) +
  xlim(3, 11) +
  ylim(0, 19000)
```

```{r}
linear.carat <- lm(x ~ carat , data=diamonds)
summary(linear.carat)$adj.r.squared
```

```{r}
ggplot(diamonds, aes(y=x, x=carat)) +
  geom_point(alpha=.5) +
  geom_smooth(method="lm")
```

```{r}
linear.carat <- lm(x ~ log(carat), data=diamonds)
summary(linear.carat)$adj.r.squared
```

```{r}
ggplot(diamonds, aes(y=x, x=carat)) +
  geom_point(alpha=.5) +
  geom_smooth(method="lm", formula=y ~ log(x))
```

```{r}
contingency <- table(diabetes, age)
contingency
```

```{r}
library(vcd)
assocstats(contingency)
```

```{r}
boxplot(mass ~ diabetes, data=PimaIndiansDiabetes2, ylab="Mass")
```

```{r}
aov <- aov(mass ~ diabetes)
summary(aov)
```

```{r}
log.model <- glm(diabetes ~ ., data=PimaIndiansDiabetes2, family=binomial)
summary(log.model)
```

```{r}
predictions <- predict(log.model , type="response")
head(predictions)
```

```{r}
fact.predictions <- rep("neg", length(diabetes))
fact.predictions[predictions > 0.5] <- "pos"
table(fact.predictions , diabetes)
```

```{r}
mean(fact.predictions == diabetes, na.rm=TRUE)
```

```{r}
fact.predictions <- rep("neg", length(diabetes))
fact.predictions[predictions > 0.6]="pos"
table(fact.predictions, diabetes)
```

```{r}
mean(fact.predictions == diabetes, na.rm=TRUE)
```

```{r}
log.model <- glm(diabetes ~ glucose, data=PimaIndiansDiabetes2, family=binomial)
```

```{r}
new.data <- data.frame(glucose=138, mass=35.5, pedigree=0.325)
predict(log.model, newdata=new.data, type="response")
```

## 5.3.6 Exercises

# 5.4 Data analysis of the flights database

```{r, echo=FALSE}
library(data.table)
flights <- fread("flights.csv")
names(flights) <- c("icao24", "callsign", "country", "last.update","last.contact", "longitude", "latitude", "altitude.baro", "ground", "velocity", "track", "vertical.rate", "sensors", "altitude.geo", "squawk", "spi", "position.source", "time")
flights[, c("last.contact", "last.update", "sensors", "altitude.geo", "squawk", "spi", "position.source"):=NULL]
flights[, altitude.baro:=ifelse(ground == TRUE , 0, altitude.baro)]
flights[, ground:=NULL]
flights[, time:=time - 1548028800]
flights <- na.omit(flights)
duplicated.rows <- duplicated(flights[, 1 : 5])
flights <- flights[!duplicated.rows]
flights <- flights[!callsign == ""]
flights <- flights[!callsign == "0000000"]
flights <- flights[!callsign == "00000000"]
flights <- flights[!grepl("TEST", callsign)]
```

```{r}
summary(flights)
```

```{r}
library(corrplot)
corrplot(cor(flights[, -c(1, 2, 3)], use="pairwise.complete.obs"))
```

```{r}
library(vcd)
assocstats(table(flights$callsign, flights$country))
```

```{r}
model.velocity <- lm(velocity ~ altitude.baro, data=flights)
summary(model.velocity)
```

```{r}
model.velocity <- lm(velocity ~ longitude + latitude + altitude.baro, data=flights)
summary(model.velocity)
```

```{r}
model.velocity <- lm(velocity ~ altitude.baro + track, data=flights)
summary(model.velocity)
```

```{r}
model.vertical <- lm(vertical.rate ~ longitude + latitude + altitude.baro + velocity, data=flights)
summary(model.vertical)
```